First steps towards a formalization of Forcing

### 2018-06-22 LSFA/intro.tex
línea que demarca el final de lo que tendría sentido.
absolutez

The current  axiomatic foundation of Set Theory is through 
first-order logic and uses the axioms by  Zermelo, Fraenkel, and
including the Axiom of Choice ($\AC$) among them. This theory is
known by the $\ZFC$ acronym. G\"odel showed that $\CH$ can't be
refuted using the $\ZFC$, unless the latter are inconsistent (we say
that \emph{$\CH$ is relatively 
  consistent with  $\ZFC$}). For a while, this result left the
possibility that one might be able show that $\ZFC\models \CH$, but in
a groundbreaking work \cite{Cohen-CH-PNAS}, Paul Cohen discovered the
technique of 
\emph{forcing} and showed that $\neg\CH$ is relatively consistent with
$\ZFC$. Forcing has since then been used for showing innumerable
independence results and perform mathematical constructions.

@@

A relative consistency proof for an axiom $A$ is then obtained by assuming that such a
$\lb M,E\rb$ exists, and constructing another $\lb M',E'\rb$
satisfying $\ZFC + A$. We single out a very special kind of models:
%
\begin{definition}\label{def:transitive-model}
  \begin{enumerate}
  \item A set $M$ (of sets) is \emph{transitive} if for all $x\in M$ and 
    $y\in x$, we have $y\in M$.
  \item $\lb M,E\rb$ is a \emph{transitive model} if $M$ is transitive
    set and  $E$ is the membership relation $\in$ restricted to
    $M$. It is \emph{countable} if $M$ is coordinable with a subset of
    $\N$; we then say that $M$ is a \emph{ctm}. 
  \end{enumerate}
\end{definition}
%
As in the last sentence, one usually uses the underlying set to refer to a transitive
model, since the relation is fixed.

@@

It is a primary need to have a means to correlate  (first-order)
properties satisfied by a model $M$ and those of a 
submodel $N\sbq M$. As a simple example on this, consider 
%% $M\defi \{\emptyset,\{\emptyset\},\{\{\emptyset\}\}\}$ and
%% $N\defi\{\{\emptyset\},\{\{\emptyset\}\}\}$, and let 
%% $\phi(x)\defi \neg\exists y (y \in x)$. It is clear that 
%% $M\models \phi(\emptyset)$; but we have
%% \[
%% M\not\models \phi(\{\emptyset\}) \quad\text{ but }\quad N\models \phi(\{\emptyset\}).
%% \]
%% There is a discrepancy between what $M$ and $N$ \emph{believe} about
%% their common element $\{\emptyset\}$.
$M\defi \{a,b,c, \{a,b\},\{a,b,c\}\}$ and
$N\defi\{a,b,\{a,b,c\}\}$, and let 
\[
\phi(x,y,z)\defi \forall w \,( w\in z \lsii w=x \o w=y).
\]
%% It is clear that  $M\models \phi(\{a,b\})$; but we have
Then we have
\[
M\not\models \phi(a,b,\{a,b,c\}) \quad\text{ but }\quad N\models \phi(a,b,\{a,b,c\}).
\]
There is a discrepancy between  $M$ and $N$ about $\{a,b,c\}$ being ``the
(unordered) pair of $a$ and $b$.'' We say that $\phi$ holds for
$a,b,\{a,b,c\}$ \emph{relative} to $N$. It is immediate to see that
$\phi$ holds  for $x,y,z$ relative to $N$ if and only if
\[
\phi^N(x,y,z)\defi \forall w.\ w\in N\limp ( w\in z \lsii w=x \o w=y)
\] 
holds. $\phi^N$ is called the \emph{relativization of $\phi$ to
  $N$}. In general, one can relativized

>>>>

The current  axiomatic foundation of Set Theory is through 
first-order logic and uses the axioms by  Zermelo, Fraenkel, and
including the Axiom of Choice ($\AC$) among them. This theory is
known by the $\ZFC$ acronym. G\"odel showed that $\CH$ can't be
refuted using the $\ZFC$, unless the latter are inconsistent (we say
that \emph{$\CH$ is relatively 
  consistent with  $\ZFC$}). For a while, this result left the
possibility that one might be able show that $\ZFC\models \CH$, but in
a groundbreaking work \cite{Cohen-CH-PNAS}, Paul Cohen discovered the
technique of 
\emph{forcing} and showed that $\neg\CH$ is relatively consistent with
$\ZFC$. Forcing has since then been used for showing innumerable
independence results and to perform mathematical constructions.

@@

A relative consistency proof for an axiom $A$ is then obtained by assuming that such a
$\lb M,E\rb$ exists, and constructing another $\lb M',E'\rb$
satisfying $\ZFC + A$. We single out a very special kind of models:
%
\begin{definition}\label{def:transitive-model}
  \begin{enumerate}
  \item A set $M$ (of sets) is \emph{transitive} if for all $x\in M$ and 
    $y\in x$, we have $y\in M$ (i.e., every element of $M$ is a subset
    of $M$).
  \item $\lb M,E\rb$ is a \emph{transitive model} if $M$ is transitive
    set and  $E$ is the membership relation $\in$ restricted to
    $M$. It is \emph{countable} if $M$ is coordinable with a subset of
    $\N$; we then say that $M$ is a \emph{ctm}. 
  \end{enumerate}
\end{definition}
%
As in the last sentence, one usually uses the underlying set to refer to a transitive
model, since the relation is fixed.

@@

It is a primary need to have a means to correlate  (first-order)
properties satisfied by a model $M$ and those of a 
submodel $N\sbq M$. As a simple example on this, consider 
%% $M\defi \{\emptyset,\{\emptyset\},\{\{\emptyset\}\}\}$ and
%% $N\defi\{\{\emptyset\},\{\{\emptyset\}\}\}$, and let 
%% $\phi(x)\defi \neg\exists y (y \in x)$. It is clear that 
%% $M\models \phi(\emptyset)$; but we have
%% \[
%% M\not\models \phi(\{\emptyset\}) \quad\text{ but }\quad N\models \phi(\{\emptyset\}).
%% \]
%% There is a discrepancy between what $M$ and $N$ \emph{believe} about
%% their common element $\{\emptyset\}$.
$M\defi \{a,b,c, \{a,b\},\{a,b,c\}\}$ and
$N\defi\{a,b,\{a,b,c\}\}$, and let 
\[
\phi(x,y,z)\defi \forall w \,( w\in z \lsii w=x \o w=y).
\]
%% It is clear that  $M\models \phi(\{a,b\})$; but we have
Then we have
\[
M\not\models \phi(a,b,\{a,b,c\}) \quad\text{ but }\quad N\models \phi(a,b,\{a,b,c\}).
\]
There is a discrepancy between  $M$ and $N$ about $\{a,b,c\}$ being ``the
(unordered) pair of $a$ and $b$.'' We say that $\phi$ holds for
$a,b,\{a,b,c\}$ \emph{relative} to $N$. It is immediate to see that
$\phi$ holds  for $x,y,z$ relative to $N$ if and only if
\[
\phi^N(x,y,z)\defi \forall w.\ w\in N\limp ( w\in z \lsii w=x \o w=y)
\] 
holds. $\phi^N$ is called the \emph{relativization of $\phi$ to
  $N$}. One can generalize this operation of relativization to the
class of all sets satisfying a first-order predicate $C$ in a
straightforward way:
\[
\phi^C(x,y,z)\defi \forall w.\ C(w)\limp ( w\in z \lsii w=x \o w=y)
\]

It can be shown elementarily that if $M$ and $N$ are transitive,
$\phi^N$ holds if and only if $\phi^M$ holds,  for $x,y,z\in N$. We
say then that $\phi$ is \emph{absolute between $N$ and $M$.}

The concept of relativization and absoluteness are central to task of
transferring truth of axioms in $M$ to  $L^M$. At this point we 
point out that although $L^M$ is a subset of $M$. 
\bigskip

\hrule
\bigskip

In general, one can relativized
